---
title: "Add title of your project"
author: "Alison Partee and Matt Gibson, Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## Directions:
1. Change "Add title of your project" on line 2 (above) to the name of your project.
2. Change "Add names of team members" on line 3 (above) to the names of team members in your project.
3. Write a <250 word Abstract describing your team's research.
4. Write a 1-2 paragraph Introduction providing some background and context for your project.
5. Set up your working enviroment (set directory, load packages) using the `/Week8-Projects` folder. 
6. Provide a brief (~1 paragraph) description of the dataset.
7. Load the dataset and process in ways necessary for any initial analyses.
8. Use well annotated R chunks to analyze data, make figures, and peform statistics needed to address your quesstions and hypotheses. 
9. Write a Discussion and Conclusion section to address the questions and hypotheses that you outlined in you outlined in your Abstract and Introuction.
10. Include citations in the Reference section.
11. Delete the text in this Directions section.
12. By Thursday March 9, 23:59, please submit the completed assignment by creating a **pull request** via GitHub. Your pull request should include this file *Project_assignment.Rmd* and the PDF output of `Knitr` (*Project_assignment.pdf*). 

## ABSTRACT

## 1) INTRODUCTION

## 2) SETUP

```{r}

rm(list=ls())
getwd()
setwd("~/GitHub/QB2017_Gibson/Week8-Projects/")


package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies')
for (package in package.list) {
  if (!require(package, character.only=T, quietly=T)){
    install.packages(package)
    library(package, character.only=T)
  }
}
package.list <- c('vegan', 'sp', 'gstat','raster', 'RgoogleMaps', 'maptools', 'rgdal',
                  'simba', 'gplots', 'rgeos', 'rgdal', "SoDA")
for (p in package.list){
  library(p, character.only = T)
  #install.packages(p)
}

```

## 3) DESCRIPTION OF DATA

## 4) LOAD THE DATA

```{r}

#setwd("c:/Users/matth/Documents/bin/QB2017_Gibson/Week8-project")

myData <- read.table("speciesdata_clean.csv", sep=",", header=T, row.names = 1)
envData <- read.table("environmentaldata(1).csv", sep=",", header=T, row.names = 1)

envData_a <- envData

##DATA TRANSFORMATIONS
envData$Topsoil.pH <- (1/(10^envData$Topsoil.pH))
envData$Subsoil.pH <- (1/(10^envData$Subsoil.pH))
envData$Topsoil.Ca <- log(envData$Topsoil.Ca+.01)
envData$Topsoil.Mg <- log(envData$Topsoil.Mg+.01)
envData$Topsoil.Mn <- log(envData$Topsoil.Mn+.01)
envData$Topsoil.Zn <- log(envData$Topsoil.Zn+.1)
envData$Topsoil.NO3 <- log(envData$Topsoil.NO3+.01)
envData$Topsoil.Ca <- log(envData$Topsoil.Ca+.01)
envData$Topsoil.N <- log(envData$Topsoil.N+.01)
envData$Topsoil.Fe <- log(envData$Topsoil.Fe+.01)
envData$Topsoil.Al <- log(envData$Topsoil.Al+.01)



```

## 5) ANALYSIS: FIGURES AND STATISTICS

```{r}
#PLOT OF SITES

library(rworldmap)
newmap <- getMap(resolution="low")
#plot(newmap)
plot(newmap, xlim = c(-10, 20), ylim = c(40, 60), asp = 1, main="Map of sites")

points(envData$Latitude, envData$Longitude, col = "red", cex = .6)

S.obs <- function(x = ""){
  rowSums(x > 0) * 1
}

C <- function(x = ""){
  1 - (sum(x == 1) / rowSums(x))
}
d<- dim(myData)

##########################################################################################
#CALCULATE RICHNESS AND COVERAGE AT ALL SITES
s <- seq(2, d[1])
richness <- S.obs(myData[1, 4:ncol(myData)])
coverage <- C(myData[1, 4:ncol(myData)])
for (i in s) {
  coverage <- rbind(coverage, C(myData[i, 4:ncol(myData)]))
  richness <- rbind(richness, S.obs(myData[i, 4:ncol(myData)]))
}
```


#####Correlations of environmental variables with richness
```{r}
##########################################################################################
#CORRELATIONS OF ENVIRONMENTAL VARIABLES WITH RICHNESS

#Remove plant chemical measurements
envData_reduced <- envData[, 1:23]
#Remove categorical variables
envData_reduced <- subset(envData_reduced, select=-c(Mangement.type, Grazing.intensity))
#Remove country and year
envData_reduced <- subset(envData_reduced, select=-c(Country, Survey.year))

bigFrame <- cbind(as.numeric(richness), envData_reduced[, 3:ncol(envData_reduced)])
colnames(bigFrame) <- c("richness", "Altitude", "Inclination", "vegetation.height", "Soil.depth", "Topsoil.pH",           "Subsoil.pH",          
 "Topsoil.Al",           "Topsoil.Ca",           "Topsoil.Fe",           "Topsoil.Mg",          
"Topsoil.Mn",           "Topsoil.Zn",           "Topsoil.NO3",          "Topsoil.NH4",         
"Topsoil.Olsen.P",      "Topsoil.C",            "Topsoil.N")
bigFrame2 <- cbind(as.numeric(richness), subset(envData, select= c(Mangement.type, Grazing.intensity, Country)))
colnames(bigFrame2) <- c("richness", "m.type", "g.type", "country")
#pairs(bigFrame)
cor1 <- cor(bigFrame)
par(mfrow=c(1,1))
library(psych)
cor2 <- corr.test(bigFrame, method="pearson", adjust="BH")

#CORRELATIONS OF ENV VARIABLES WITH RICHNESS
cor2$r[,1]

```

#####Plots of richness vs. various environmental variables
```{r, fig.width=8, fig.height=8}
##########################################################################################
#PLOT OF ENV VARIABLES VS RICHNESS

par(mfrow=c(2,2))

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.pH), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.pH, bigFrame$richness, ylab = "Richness", xlab = "[H+]", main="Richness vs. \n Hydrogen Ion Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = .00015, y = 50, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Ca), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Ca, bigFrame$richness, ylab = "Richness", xlab = "Calcium", main = "Richness vs.\n Calcium Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 1, y = 50, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Mg), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Mg, bigFrame$richness, ylab = "Richness", xlab = "Magnesium",main = "Richness vs. \n Magnesium Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 6, y = 50, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Mn), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Mn, bigFrame$richness, ylab = "Richness", xlab = "Manganese",main = "Richness vs. \n Manganese Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = .00015, y = 55, labels = mylabel, cex=.9)
abline(model1)

par(mfrow=c(2,2))

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Zn), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Zn, bigFrame$richness, ylab = "Richness", xlab = "Zinc",main = "Richness vs. \n Zinc Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 3, y = 55, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.NO3), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.NO3, bigFrame$richness, ylab = "Richness", xlab = "Nitrates",main = "Richness vs. \n Nitrate Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = -2, y = 54, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.N), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.N, bigFrame$richness, ylab = "Richness", xlab = "Nitrogen",main = "Richness vs. \n Nitrogen Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 2, y = 55, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Fe), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Fe, bigFrame$richness, ylab = "Richness", xlab = "Iron",main = "Richness vs. \n Iron Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 2, y = 55, labels = mylabel, cex=.9)
abline(model1)

par(mfrow=c(2,2))

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Al), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Al, bigFrame$richness, ylab = "Richness", xlab = "Aluminium",main = "Richness vs. \n Aluminium Concentration")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 6, y = 53, labels = mylabel, cex=.9)
abline(model1)

model1 <- lm(as.numeric(richness) ~ na.omit(Topsoil.Olsen.P), data=bigFrame)
print(s <- summary(model1))
plot(bigFrame$Topsoil.Olsen.P, bigFrame$richness, ylab = "Richness", xlab = "Olsen Phos.",main = "Richness vs. \n Olsen Phosphorus")
mylabel = bquote(italic(P) == .(format(s$coefficients[2,4], digits = 3)))
text(x = 60, y = 50, labels = mylabel, cex=.9)
abline(model1)


```

#####Plots of mean chemical composition by country
```{r, fig.width=8, fig.height=8}

##########################################################################################
#PLOT OF ENV VARIABLES VS RICHNESS




plot.new()
#envData_a <- envData
envData_a <- na.omit(envData_a)
envData_a$region <- as.factor(substr(rownames(envData_a),1,2))
#REG <- factor(envData_a$regions, levels=c('UK', 'SE', 'GE', 'DK', 'NL', 'BE', 'NW', 'FR'))

#means_topPH <- tapply(envData_a$Topsoil.pH, envData_a$region, mean)
#means_subPH <- tapply(envData_a$Subsoil.pH, envData_a$region, mean)
#means_topAl <- tapply(envData_a$Topsoil.Al, envData_a$region, mean)
#means_topCa <- tapply(envData_a$Topsoil.Ca, envData_a$region, mean)
#means_topFe <- tapply(envData_a$Topsoil.Fe, envData_a$region, mean)
#means_topMg <- tapply(envData_a$Topsoil.Mg, envData_a$region, mean)
#means_topMn <- tapply(envData_a$Topsoil.Mn, envData_a$region, mean)
#means_topZn <- tapply(envData_a$Topsoil.Zn, envData_a$region, mean)
#means_topNO3 <- tapply(envData_a$Topsoil.NO3, envData_a$region, mean)
#means_topNH4 <- tapply(envData_a$Topsoil.NH4, envData_a$region, mean)
#means_topOP <- tapply(envData_a$Topsoil.Olsen.P, envData_a$region, mean)
#means_topC <- tapply(envData_a$Topsoil.C, envData_a$region, mean)
#means_topN <- tapply(envData_a$Topsoil.N, envData_a$region, mean)
#means_topP <- tapply(envData_a$Topsoil.P, envData_a$region, mean)

sem <- function(x){
  sd(na.omit(x))/sqrt(length(na.omit(x)))
}

#zp.topPH <- tapply(envData_a$Topsoil.pH, envData_a$region, sem)
#zp.subPH <- tapply(envData_a$Subsoil.pH, envData_a$region, sem)
#zp.topAl <- tapply(envData_a$Topsoil.Al, envData_a$region, sem)
#zp.topCa <- tapply(envData_a$Topsoil.Ca, envData_a$region, sem)
#zp.topFe <- tapply(envData_a$Topsoil.Fe, envData_a$region, sem)
#zp.topMg <- tapply(envData_a$Topsoil.Mg, envData_a$region, sem)
#zp.topMn <- tapply(envData_a$Topsoil.Mn, envData_a$region, sem)
#zp.topZn <- tapply(envData_a$Topsoil.Zn, envData_a$region, sem)
#zp.topNO3 <- tapply(envData_a$Topsoil.NO3, envData_a$region, sem)
#zp.topNH4 <- tapply(envData_a$Topsoil.NH4, envData_a$region, sem)
#zp.topOP <- tapply(envData_a$Topsoil.Olsen.P, envData_a$region, sem)
#zp.topC <- tapply(envData_a$Topsoil.C, envData_a$region, sem)
#zp.topN <- tapply(envData_a$Topsoil.N, envData_a$region, sem)

variables <- c('Topsoil.pH', "Subsoil.pH", "Topsoil.Al", 'Topsoil.Ca', 'Topsoil.Fe', 'Topsoil.Mg', "Topsoil.Mn", 'Topsoil.Zn', 'Topsoil.NO3', 'Topsoil.NH4', 
               'Topsoil.Olsen.P', 'Topsoil.C', 'Topsoil.N')


par(mfrow=c(4,4))

i <- 1
for (x in 11:23){
  #print(variables[i])
  mean1 <- tapply(envData_a[,x], envData_a$region, mean)
  #print(mean1)
  sem1 <- tapply(envData_a[,x], envData_a$region, sem)
  bp <- barplot(mean1,ylim = c(0, round(max(mean1)+5, digits = 0)),
              pch = 15, cex = 1.25, las = 1, cex.lab = 1, cex.axis = 1, cex.names = .8,
              xlab = "Region", ylab = variables[i], las = 2
              )
  
  arrows(x0 = bp, y0 = mean1, y1 = mean1 - sem1, angle = 90,
       length = 0.05, lwd=.8)
  arrows(x0 = bp, y0 = mean1, y1 = mean1 + sem1, angle = 90,
       length = 0.05, lwd=.8)
  i <- i + 1
  
}

```


#####Ordination of taxonomic diversity
```{r, fig.width=8, fig.height=8}
par(mfrow=c(1,1))

myData <- myData[, 4:ncol(myData)]
spec.bray <- vegdist(myData, method = "bray")

spec.pcoa <- cmdscale(spec.bray, eig=T, k=3)
explainvar1 <- round(spec.pcoa$eig[1] / sum(spec.pcoa$eig), 3) * 100
explainvar2 <- round(spec.pcoa$eig[2] / sum(spec.pcoa$eig), 3) * 100
explainvar3 <- round(spec.pcoa$eig[3] / sum(spec.pcoa$eig), 3) * 100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

#Variance explained by first axis
explainvar1
#variance explained by second axis
explainvar2
#variance explained by third axis
explainvar3
sum.eig

par(mar = c(5, 5, 1, 2) + 0.1)

plot(spec.pcoa$points[ ,1], spec.pcoa$points[ ,2], ylim = c(-0.6, 0.45),
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
     main = "Taxonomic Ordination",
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = F)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)




data2 <- as.data.frame(spec.pcoa$points)
data2$cun <- substr(row.names(data2),1,2)
fr <- data2[which(data2$cun == "FR"),]
nw <- data2[which(data2$cun == "NW"),]
be <- data2[which(data2$cun == "BE"),]
nl <- data2[which(data2$cun == "NL"),]
dk <- data2[which(data2$cun == "DK"),]
ge <- data2[which(data2$cun == "GE"),]
se <- data2[which(data2$cun == "SE"),]
uk <- data2[which(data2$cun == "UK"),]


points(fr[,1], fr[,2] ,pch=19, cex=2, bg="red", col="red")
points(nw[,1], nw[,2],pch=19, cex=2, bg="blue", col="blue")
points(be[,1], be[,2],pch=19, cex=2, bg="chartreuse2", col="chartreuse2")
points(nl[,1], nl[,2],pch=19, cex=2, bg="darkgoldenrod1", col="darkgoldenrod1")
points(dk[,1], dk[,2],pch=19, cex=2, bg="darkgray", col="darkgray")
points(ge[,1], ge[,2],pch=19, cex=2, bg="darkorange1", col="darkorange1")
points(se[,1], se[,2],pch=19, cex=2, bg="darksalmon", col="darksalmon")
points(uk[,1], uk[,2],pch=19, cex=2, bg="darkviolet", col="darkviolet")

legend(.4,-.3,c("FR","NW", "BE", "NL", "DK", "GE", "SE", "UK"), cex = .7, col=c("red", "blue", "chartreuse2", "darkgoldenrod1",
                                                                   "darkgray", "darkorange1", "darksalmon", "darkviolet"), fill =c("red", "blue", "chartreuse2", "darkgoldenrod1","darkgray", "darkorange1", "darksalmon", "darkviolet") )
#I plotted just the country code...full site names were too long
#(spec.pcoa$points[ ,1], spec.pcoa$points[ ,2],
#       pch = 19, cex = 3, bg = "gray", col = "gray")
#text(spec.pcoa$points[ ,1], spec.pcoa$points[ ,2],
#     labels = substr(row.names(spec.pcoa$points),1,2))

specREL <- myData
  for(i in 1:nrow(myData)){
    specREL[i, ] = myData[i, ]/ sum(myData[i, ])
  }

spec.pcoa <- add.spec.scores(spec.pcoa, specREL, method = "pcoa.scores")
par(mfrow=c(1,1))
#Is just a mess.........
#text(spec.pcoa$cproj[ ,1], spec.pcoa$cproj[ ,2],
#     labels = row.names(spec.pcoa$cproj), col = "black")


spe.corr <- add.spec.scores(spec.pcoa, specREL, method = "cor.scores")$cproj
corrcut <- 0.8
imp.spp <- spe.corr[abs(spe.corr[, 1]) >= corrcut | abs(spe.corr[, 2]) >= corrcut, ]

#As expected, this takes quite a long time...

fit <- envfit(spec.pcoa, specREL, perm = 999)
fit

```



#####Constrained Ordination
```{r, fig.width=8, fig.height=8}
#Isolating the environmental chemical data (removing categorical variables and other plant chemical measurements)
env.chem <- na.omit(envData_reduced[, 7:19])
env.chem <- as.matrix(env.chem)
#is.na(env.chem) <- do.call(cbind,lapply(env.chem, is.infinite))

###REMOVING SITES WITH MISSING DATA
myData <- myData[rownames(myData) != "GE712", ]
myData <- myData[rownames(myData) != "GE713", ]
myData <- myData[rownames(myData) != "GE715", ]

spec.db <- vegdist(myData, method = "bray", diag=T)
eu.dbrda <- dbrda(spec.db ~ ., as.data.frame(env.chem), na.action=na.omit)
#ordiplot(eu.dbrda)
eu.dbrda.mod0 <- dbrda(spec.db ~ 1, as.data.frame(env.chem), na.action=na.omit)

#ordiplot(eu.dbrda.mod0)
eu.dbrda.mod1 <- dbrda(spec.db ~ ., as.data.frame(env.chem), na.action=na.omit)
#Model selection
eu.dbrda <- ordiR2step(eu.dbrda.mod0, eu.dbrda.mod1, perm.max= 200)

eu.dbrda$call
eu.dbrda$anova
ordiplot(eu.dbrda)
par(mfrow=c(1,1))

#3. use a permutation test to determine the significance of the constrained analysis,
permutest(eu.dbrda, permutations = 999)
#4 use a permutation test to determine the correlation of each environmental factor on the constrained axes,
envfit(eu.dbrda, env.chem[,c(1,3,4,5,6,7,9,12,13)], perm = 999)

#5
dbrda.explainvar1 <- round(eu.dbrda$CCA$eig[1] /
                             sum(c(eu.dbrda$CCA$eig, eu.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(eu.dbrda$CCA$eig[2] /
                             sum(c(eu.dbrda$CCA$eig, eu.dbrda$CA$eig)), 3) * 100
dbrda.explainvar1
dbrda.explainvar2
#6
par(mar = c(5, 5, 4, 4) + 0.1)

plot(scores(eu.dbrda, display = "wa"), xlim = c(-1.5, 2.2), ylim = c(-2.4, 2.5),
xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)", sep = ""),
ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

data <- as.data.frame(scores(eu.dbrda)$sites)
data$cun <- substr(row.names(data),1,2)
fr <- data[which(data$cun == "FR"),]
nw <- data[which(data$cun == "NW"),]
be <- data[which(data$cun == "BE"),]
nl <- data[which(data$cun == "NL"),]
dk <- data[which(data$cun == "DK"),]
ge <- data[which(data$cun == "GE"),]
se <- data[which(data$cun == "SE"),]
uk <- data[which(data$cun == "UK"),]

points(fr$dbRDA1, fr$dbRDA2,pch=19, cex=2, bg="red", col="red")
points(nw$dbRDA1, nw$dbRDA2,pch=19, cex=2, bg="blue", col="blue")
points(be$dbRDA1, be$dbRDA2,pch=19, cex=2, bg="chartreuse2", col="chartreuse2")
points(nl$dbRDA1, nl$dbRDA2,pch=19, cex=2, bg="darkgoldenrod1", col="darkgoldenrod1")
points(dk$dbRDA1, dk$dbRDA2,pch=19, cex=2, bg="darkgray", col="darkgray")
points(ge$dbRDA1, ge$dbRDA2,pch=19, cex=2, bg="darkorange1", col="darkorange1")
points(se$dbRDA1, se$dbRDA2,pch=19, cex=2, bg="darksalmon", col="darksalmon")
points(uk$dbRDA1, uk$dbRDA2,pch=19, cex=2, bg="darkviolet", col="darkviolet")

legend(1.5,2.3,c("FR","NW", "BE", "NL", "DK", "GE", "SE", "UK"), cex = .7, col=c("red", "blue", "chartreuse2", "darkgoldenrod1",
                                                                   "darkgray", "darkorange1", "darksalmon", "darkviolet"), fill =c("red", "blue", "chartreuse2", "darkgoldenrod1","darkgray", "darkorange1", "darksalmon", "darkviolet") )
#text(scores(eu.dbrda, display = "wa"),
 # labels = substr(row.names(scores(eu.dbrda, display = "wa")),1,2))

vectors <- scores(eu.dbrda, display = "bp")

#row.names(vectors) <- c("pH", "har", "pho", "nit", "amm", "oxy", "bdo")

arrows(0, 0, vectors[,1], vectors[, 2],
  lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[, 2], pos = 3,
  labels = row.names(vectors))

axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))

par(mfrow=c(1,1))

```

####Variance Partitioning
```{r, fig.width=8, fig.height=8}
#Remove plant measurements
envData_reduced <- envData[, 1:23]
#Remove categorical variables
envData_reduced <- subset(envData_reduced, select=-c(Mangement.type, Grazing.intensity))
#Remove country and year
envData_reduced <- subset(envData_reduced, select=-c(Country, Survey.year))
envData_reduced <- envData_reduced[rownames(envData_reduced) != "GE712", ]
envData_reduced <- envData_reduced[rownames(envData_reduced) != "GE713", ]
envData_reduced <- envData_reduced[rownames(envData_reduced) != "GE715", ]
#Remove unneeded data
myData <- myData[, 4:ncol(myData)]
XY_soda <- as.data.frame(geoXY(envData_reduced$Latitude, envData_reduced$Longitude))
#XY_soda <- cbind(row.names((envData_reduced)), XY_soda)
#colnames(XY_soda) <- c("site", "X", "Y")
#Begin DISTANCE DECAY CODE
xy <- data.frame(site.name = row.names(envData_reduced), lats = envData_reduced$Latitude, lons = envData_reduced$Longitude)
#coordinates(xy) <- ~lats+lons
comm.dist <- 1 - vegdist(myData)
#proj4string(xy) <- CRS("+proj=longlat +datum=NAD83")
#UTM <- spTransform(xy, CRS("+proj=utm + zone=51 ellps=WGS84"))
#UTM <- as.data.frame(UTM)
xy$lats_utm <- XY_soda$Y
xy$lons_utm <- XY_soda$X

lats <- XY_soda$Y
lons <- XY_soda$X

# 3) Calculate geographic distance between plots and assign to the variable 'coord.dist'
coord.dist <- dist(as.matrix(lats, lons))

par(mfrow=c(1,1))

eu.dbrda$anova

env.mod <- model.matrix(~ Topsoil.pH + Topsoil.Ca + Topsoil.Mn + Topsoil.C + Topsoil.NO3 + Topsoil.N + Topsoil.Zn + Topsoil.Al , as.data.frame(env.chem))[,-1]
rs <- rowSums(myData)/sum(myData)
xy <- as.data.frame(xy)
coord.mat <- as.matrix(xy[,4:5])

doubs.pcnmw <- pcnm(dist(coord.mat), w = rs, dist.ret = T)
doubs.pcnmw$values > 0

doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(spec.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(spec.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

plot(step.pcnm)

step.pcnm$anova

space.mod <- model.matrix(~ PCNM1 + PCNM3 + PCNM2 + PCNM6 + PCNM4 + PCNM9 + PCNM7 + PCNM11 + PCNM8 + PCNM5, doubs.space)[,-1]

doubs.total.env <- dbrda(spec.db ~ env.mod)
doubs.total.space <- dbrda(spec.db ~ space.mod)
doubs.env.cond.space <- dbrda(spec.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(spec.db ~ space.mod + Condition(space.mod))

permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

doubs.varpart <- varpart(spec.db, env.mod, space.mod)
doubs.varpart

par(mar = c(2,2,2,2))
plot(doubs.varpart, col ="red", cex = 1.5)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation partitioning of \nEuropean Plant Diversity", side = 3, line = -3)

```

## 6) DISCUSSION AND CONCLUSION

## 7) REFERENCES
